\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=0.8in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage[colorlinks=true,linkcolor=black,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{fancyhdr}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.3em}
\setlength{\topsep}{1pt}\setlength{\itemsep}{0pt}\setlength{\parsep}{0pt}
\pagestyle{fancy}\fancyhf{}\rhead{\small Slippay --- Governance for the Agent Economy}\lhead{\small Extended Abstract \textbar\ NEAR X}\cfoot{\thepage}
\renewcommand{\headrulewidth}{0.2pt}

\title{\textbf{Slippay: A Portable Governance Layer for the Agent Economy}\\[2pt]
\large Extended Abstract for NEAR X}
\author{Manuel Guilherme Galmanus \;\textbar\; Bluewave AI \;\textbar\; \texttt{manuel@bluewaveai.online}}
\date{June 26, 2026}

\begin{document}
\maketitle
\thispagestyle{fancy}
\vspace{-1.2em}

\section*{The thesis in one line}
Agent \emph{payment} is solved; agent \emph{governance} is not --- and governance is the layer
NEAR's ``Currency of Agents'' needs to be trusted at scale. Slippay is a working, chain-agnostic
governance primitive for it, with a transaction hash behind the claim.

\section*{What is already solved (so we don't reinvent it)}
Over HTTP~402, autonomous agents already pay each other in stablecoins: Coinbase/Cloudflare's
\textbf{x402} has settled 161M+ machine payments ($\approx$\$43.5M); Google's \textbf{AP2}
standardizes human-approved spending \emph{mandates}. NEAR itself routes agent goals across
chains via \textbf{Intents}, matches work in the \textbf{AI Agent Market}, and hardens execution
in \textbf{TEEs} (IronClaw / Intel~TDX). The rails and the routing exist.

\section*{What is NOT solved --- the governance gap}
When \emph{both} sides of a transaction are autonomous and no human clicked, three questions stay
open, and they are the ones that decide whether an agent economy is investable and compliant:
\begin{enumerate}
  \item \textbf{Holder / titularity.} Who is the legal holder of the transaction? Multi-agent
  liability law is openly unsettled (Berkeley Tech.\ Law J., Columbia CLS, 2026).
  \item \textbf{Integrity.} Is the paying agent still uncompromised, or hijacked into an
  unauthorized payment (llms.txt / prompt injection)?
  \item \textbf{Bound provenance.} Is the agent's spending limit \emph{proven}, or just a
  runtime check a compromised server could skip?
\end{enumerate}
These are one problem: the \emph{self-governance} of an autonomous payer.

\section*{Slippay's answer (and why it is proven, not a slide)}
We treat payment authorization as a \textbf{metacognitive monitor--control loop}: a payment
settles only when a monitor confirms the agent is inside a human-rooted, signed surface, and that
verdict is itself a portable, checkable artifact. Three mechanisms, all \textbf{non-custodial}
(the agent's own wallet signs; no platform holds funds):
\begin{itemize}
  \item \textbf{Agent-Integrity Attestation (AIA).} An \mbox{ed25519} verdict ---``is this agent
  inside its committed surface?''--- bound to the payment, verifiable \emph{on-chain and offline}
  by anyone, fail-closed. Not a platform-locked guardrail.
  \item \textbf{AXL proof-carrying bound.} The spending limit compiles to an SMT (z3) certificate
  anyone can re-verify offline: the agent \emph{cannot} exceed the cap, provably.
  \item \textbf{Human principal at the root.} One signed mandate is the authorization; consent
  exists without a per-transaction click (aligned with UETA/E-SIGN deployer-as-principal and
  AP2's mandate chains).
\end{itemize}

\textbf{Demonstrated end-to-end on Stellar mainnet} --- a buyer agent paid a seller agent inside
its mandate, with no human in the loop, and received the good:
\begin{center}\small
\begin{tabular}{@{}lll@{}}
\toprule
Autonomous A2A payment & Stellar mainnet & \texttt{34eb71b1\ldots ea13b} (ledger 63{,}206{,}144)\\
Full loop incl.\ delivery & Stellar testnet & \texttt{61db50c6\ldots 1d55f1} (ledger 3{,}291{,}957)\\
\bottomrule
\end{tabular}
\end{center}

\section*{Why this fits NEAR specifically}
NEAR is building the \emph{cross-chain} agent economy; Slippay's verdict is \textbf{chain-agnostic
by construction} --- the same signed bytes verify on a Soroban contract and off-chain on any
stack, so it travels with the agent across the chains NEAR Intents route over.
\begin{itemize}
  \item \textbf{Complements NEAR's TEE bet, doesn't compete with it.} TEEs answer ``trust the
  execution environment.'' Slippay answers ``verify the claim'' --- a portable cryptographic
  attestation + a proven bound that hold \emph{without} trusting the host. Combined: a TEE
  attests, Slippay's portable verdict carries that trust cross-chain and gates settlement.
  \item \textbf{The governance layer the Agent Market needs.} Escrowed, autonomous task-payment
  becomes safe when each payout is integrity-gated and mandate-bounded --- exactly Slippay's gate.
  \item \textbf{Regulatory tailwind.} Brazil's DeCripto (IN~2{,}291, Jul~2026) and the OECD CARF
  (70+ countries) oblige identification + reporting while carving out \emph{self-custody}. A
  cross-chain agent economy faces this globally; the payments that survive are non-custodial,
  mandate-bounded and auditable --- Slippay's exact shape.
\end{itemize}

\section*{The ask}
A conversation about making this the \textbf{open governance layer for NEAR's agent economy}: a
pilot wiring the AIA + proof-carrying mandate into a NEAR Intents / Agent-Market flow, and/or an
ecosystem grant to harden the chain-agnostic attestation across NEAR's cross-chain execution. We
bring a proven primitive (on-chain, today) and the research behind it; NEAR brings the agent
economy that needs it.

\medskip\hrule\medskip
\small
\textbf{Applied paper:} \emph{Slippay --- Metacognitive Governance for Autonomous Agent Payments:
Non-Custodial, Integrity-Gated, Proof-Carrying Settlement on Stellar.}
\url{https://github.com/Galmanus/slippay/blob/main/docs/paper/slippay.pdf}\\
\textbf{Foundational paper:} \emph{Metacognitive Engineering as a Security Surface: Introspective
Vulnerabilities in Large Language Model Alignment} (41~pp).
\url{https://github.com/Galmanus/slippay/blob/main/docs/paper/metacognitive-security-surface.pdf}\\
\textbf{Code (open):} \url{https://github.com/Galmanus/slippay} \;\textbar\;
\textbf{Live:} \url{https://app.slippay.cc}\\
\textbf{Honest positioning:} agent payment is not novel (x402/AP2 cited); the contribution is the
integrity-gated, proof-carrying, metacognitive synthesis and a publicly verifiable artifact.

\end{document}
